---
title: "GutMicrobiomeMetabolomeCompendium"
author: "Nalin Arora, Shivangi Verma"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{GutMicrobiomeMetabolomeCompendium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`GutMicrobiomeMetabolomeCompendium` provides curated study-level tables for:

- metadata
- species profiles
- metabolome profiles

The package includes helper functions to list studies, inspect available
modalities, and load one or more studies into R.

## Installation

From Bioconductor:

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("GutMicrobiomeMetabolomeCompendium")
```

Development version from GitHub:

```{r eval = FALSE}
install.packages("remotes")
remotes::install_github(
    "Shivi-Verma29/Microbiome_Metabolome_Curated_Studies"
)
```

## Load the package

```{r}
library(GutMicrobiomeMetabolomeCompendium)
```

## Discover available studies

```{r}
studies <- GMMC_studies()
length(studies)
head(studies)
```

## Inspect the study index

```{r}
idx <- GMMC_study_index()
head(idx[, c("study", "has_metadata", "has_species", "has_metabolome")])
```

## Load one study

```{r}
study_id <- idx$study[[1]]
available <- GMMC_available_modalities(study_id)
available

one_study <- GMMC_load_study(study_id)
names(one_study)
```

## Load one modality

```{r}
meta <- GMMC_load_modality(study_id, "metadata")
dim(meta)
head(meta[, seq_len(min(6, ncol(meta)))])
```

## Load multiple studies and modalities

```{r}
selected_studies <- studies[seq_len(min(2, length(studies)))]
multi <- GutMicrobiomeMetabolomeCompendium(
    studies = selected_studies,
    modalities = c("metadata", "metabolome")
)

names(multi)
names(multi[[1]])
```

## Session information

```{r}
sessionInfo()
```
