---
title: "HuMMANet"
author: "Nalin Arora, Shivangi Verma"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{HuMMANet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`HuMMANet` is intended as a Bioconductor data resource for paired human
microbiome-metabolome studies. It provides a unified interface to curated
study-level species profiles, metabolite profiles, matched sample metadata, and
study-specific annotation resources, with study data accessible through
`ExperimentHub`.

The central motivation for the package is to provide a reusable resource for
paired study species and metabolite data together with supporting annotations,
so that users can work with these studies in a consistent and reproducible way.
Rather than manually assembling heterogeneous files from individual studies,
users can access matched tables and annotation resources through a common API in
the Bioconductor ecosystem.

Each study bundle can include the following modalities:

- `MetadataProfile`: sample-level study metadata, including sample identifiers,
  subject or participant annotations, clinical covariates, longitudinal labels,
  body site information, and other variables needed to interpret the study.
- `taxaAbundanceProfile`: the curated microbial abundance table for a study,
  aligned to the study samples and representing taxa or related microbial
  features.
- `OriginalMetaboliteProfile`: the curated metabolomics matrix prior to
  HuMMANet harmonization, preserving the original study-specific metabolite
  feature space.
- `harmonizedMetaboliteProfile`: the HuMMANet-processed metabolomics matrix
  after harmonization, designed to support more consistent comparisons across
  studies.
- `metaboliteIdentifierMapping`: metabolite identifier mapping tables that link
  HuMMANet identifiers to external resources such as HMDB, PubChem, KEGG,
  ChEBI, structural descriptors, and standardized metabolite names.
- `microbialProducerAnnotations`: curated annotations linking metabolites to
  potential microbial producers or species-level producer evidence.
- `metaboliteDiseaseAssociations`: curated disease associations for metabolites
  represented in a study.
- `metabolitePathwayAnnotations`: curated pathway annotations for study
  metabolites.
- `drugBankSimilarityMatches`: DrugBank similarity matches for study
  metabolites.
- `drugCentralSimilarityMatches`: DrugCentral similarity matches for study
  metabolites.

The package is structured as a lightweight `ExperimentHub` interface package.
Large study bundles are published outside the source tarball, while the package
ships accessor functions, metadata, and documentation.

## Installation

Install from Bioconductor:

```{r eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("HuMMANet")
```

Install the development version from GitHub:

```{r eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_github("ShivangiV-IIITD/HuMMANet")
```

## Load the package

```{r}
library(HuMMANet)
```

## Connect to ExperimentHub

```{r eval = FALSE}
hub <- HuMMANetHub()
hub
```

## Discover available studies

```{r}
studies <- HuMMANet_studies()
length(studies)
head(studies)
```

## Inspect the study index

```{r}
idx <- HuMMANet_study_index()
head(
  idx[, c(
    "study",
    "has_MetadataProfile",
    "has_taxaAbundanceProfile",
    "has_OriginalMetaboliteProfile",
    "has_harmonizedMetaboliteProfile"
  )]
)
```

## Load one study

```{r}
study_id <- idx$study[[1]]
available <- HuMMANet_available_modalities(study_id)
available

one_study <- HuMMANet_load_study(study_id)
names(one_study)
```

## Load one modality

```{r}
metadata_profile <- HuMMANet_load_modality(study_id, "MetadataProfile")
dim(metadata_profile)
head(metadata_profile[, seq_len(min(6, ncol(metadata_profile)))])
```

## Load the harmonized metabolite profile

```{r eval = FALSE}
harmonized_profile <- HuMMANet_load_modality(
  study_id,
  "harmonizedMetaboliteProfile"
)
dim(harmonized_profile)
```

## Load multiple studies and modalities

```{r}
selected_studies <- studies[seq_len(min(2, length(studies)))]
multi <- HuMMANet(
    studies = selected_studies,
    modalities = c("MetadataProfile", "OriginalMetaboliteProfile")
)

names(multi)
names(multi[[1]])
```

## Session information

```{r}
sessionInfo()
```
