---
title: "HuMMANet"
author: "Nalin Arora, Shivangi Verma"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{HuMMANet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`HuMMANet` provides curated study-level tables for:

- metadata
- species profiles
- metabolome profiles

The package is structured as a lightweight `ExperimentHub` interface package.
Large study bundles are published outside the source tarball, while the package
ships accessor functions, metadata, documentation, and hub-preparation scripts.

## Installation

Install from Bioconductor:

```{r eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("HuMMANet")
```

## Load the package

```{r}
library(HuMMANet)
```

## Connect to ExperimentHub

```{r eval = FALSE}
hub <- HuMMANetHub()
hub
```

## Discover available studies

```{r}
studies <- HuMMANet_studies()
length(studies)
head(studies)
```

## Inspect the study index

```{r}
idx <- HuMMANet_study_index()
head(idx[, c("study", "has_metadata", "has_species", "has_metabolome")])
```

## Load one study

```{r}
study_id <- idx$study[[1]]
available <- HuMMANet_available_modalities(study_id)
available

one_study <- HuMMANet_load_study(study_id)
names(one_study)
```

## Load one modality

```{r}
meta <- HuMMANet_load_modality(study_id, "metadata")
dim(meta)
head(meta[, seq_len(min(6, ncol(meta)))])
```

## Load multiple studies and modalities

```{r}
selected_studies <- studies[seq_len(min(2, length(studies)))]
multi <- HuMMANet(
    studies = selected_studies,
    modalities = c("metadata", "metabolome")
)

names(multi)
names(multi[[1]])
```

## Local hub-resource preparation

For package development, the repository keeps local csv files outside the built
source package. To generate `.rds` study bundles plus a metadata template for
submission to `ExperimentHub`, run:

```{r eval = FALSE}
source("inst/scripts/make-experimenthub-resources.R")
build_hummanet_experimenthub_resources()
```

## Session information

```{r}
sessionInfo()
```
